Modeling site-specific amino-acid preferences deepens phylogenetic estimates of the divergence of viral proteins.

### 2018-03-13 manuscript/manuscript.tex
changed "tree height" to "tree diameter" per Erick Matsen

Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree. 
\ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
The tree's deepest branches are shortest when they are optimized by the GY94 model, which lacks both $\Gamma\omega$ and site-specific amino-acid preferences.
Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
Specifically, the tree's total height (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree height} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
The total height is \jdbcomment{8\% needs to be updated for tree height} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called ``tree's total height'' here.
So can you update \ref{supptab:tree_lengths} and the preceding numbers to use that measure instead?}

The deepening of branch lengths that results from the $\Gamma\omega$ and site-specific amino-acid preference approaches to modeling purifying selection are largely independent.
This can be seen by examining the ExpCM+$\Gamma\omega$ models, which combine $\Gamma\omega$ rate variation with site-specific amino-acid preferences.
As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree height} of the total tree height extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).

@@

\begin{supptable}[t!]
\caption{\label{supptab:tree_lengths}
{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
The branch lengths are in average codon substitutions per site. 
\jdbcomment{Change to be total tree height, which is distance between two most distant tip nodes.
Also, to make it easier to relate back to the percentage increases in the text, make 8 rows, one for each model and get rid of the ``rate variation'' column. 
Then add a new column that gives \% height relative to GY94 (the entry would just be 100\% for GY94).
So this gives 3 columns: model, total tree height, \% of GY94 tree height}
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & no rate variation & {\shortstack{$\Gamma\omega$ rate\\ variation}} \\ \hline
         GY94 &53.05 & 72.28\\
         ExpCM(H1) & 57.32 & 81.31\\
         ExpCM(H3) & 57.77 & 78.98\\
         ExpCM(H1 + H3 avg) & 62.98 & 83.96 \\
      \end{tabular}
\end{supptable}

>>>>

Importantly, models that account for purifying selection via either $\Gamma\omega$ rate variation or a site-specific amino-acid preferences do not just exhibit better fit---they also estimate longer deep branches on the HA tree. 
\ref{fig:empirical_trees} shows the branch lengths optimized by each model on a common scale.
The tree's deepest branches are shortest when they are optimized by the GY94 model, which lacks both $\Gamma\omega$ and site-specific amino-acid preferences.
Adding either $\Gamma\omega$ rate variation or site-specific amino-acid preferences increases the length of the deep branches.
Specifically, the tree's diameter (distance from the two most divergent tips) is \jdbcomment{36\% needs to be updated for tree diameter} longer for the GY94+$\Gamma\omega$ model than for the GY94 model (\ref{supptab:tree_lengths}).
The tree diameter is \jdbcomment{8\% needs to be updated for tree diameter} longer for ExpCM's informed by H1 or H3 amino-acid preferences than for GY94, and \jdbcomment{?} longer for the ExpCM informed by the average of the H1 and H3 preferences (\ref{supptab:tree_lengths}).
\jdbcomment{I am pretty sure that the standard quantity used in this context is not the average or total branch length, but rather the distance between the two most distant nodes on the tree, which is what I have called the ``tree diameter'' here.
I checked with Erick Matsen, and he said that ``tree diameter'' is the correct term here.
So can you update \ref{supptab:tree_lengths} and the preceding numbers to use that measure instead?}

The deepening of branch lengths that results from the $\Gamma\omega$ and site-specific amino-acid preference approaches to modeling purifying selection are largely independent.
This can be seen by examining the ExpCM+$\Gamma\omega$ models, which combine $\Gamma\omega$ rate variation with site-specific amino-acid preferences.
As shown in \ref{fig:empirical_trees}, these ExpCM+$\Gamma\omega$ models estimate longer branches than models with just $\Gamma\omega$ rate variation (GY94+$\Gamma\omega$) or just site-specific amino-acid preferences (ExpCM's).
The near independence of these effects is quantified in \ref{supptab:tree_lengths}, which shows that \jdbcomment{96\%, needs to be fixed if using tree diameter} of the tree diameter extension of ExpCM(H1+H3)+$\Gamma\omega$ versus can be explained by simply adding the extension from incorporating $\Gamma\omega$ (GY94+$\Gamma\omega$ versus GY94) to the extension from incorporating site-specific amino-acid preferences (ExpCM(H1+H3) versus GY94).

@@

\begin{supptable}[t!]
\caption{\label{supptab:tree_lengths}
{\bf Total branch length for the trees in \ref{fig:empirical_trees}.}
The branch lengths are in average codon substitutions per site. 
\jdbcomment{Change to be tree diameter, which is distance between two most distant tip nodes.
Also, to make it easier to relate back to the percentage increases in the text, make 8 rows, one for each model and get rid of the ``rate variation'' column. 
Then add a new column that gives \% diameter relative to GY94 (the entry would just be 100\% for GY94).
So this gives 3 columns: model, tree diameter, \% of GY94 tree diameter}
} 
     \begin{tabular}{cccccccccc}
        \hline
         Model & no rate variation & {\shortstack{$\Gamma\omega$ rate\\ variation}} \\ \hline
         GY94 &53.05 & 72.28\\
         ExpCM(H1) & 57.32 & 81.31\\
         ExpCM(H3) & 57.77 & 78.98\\
         ExpCM(H1 + H3 avg) & 62.98 & 83.96 \\
      \end{tabular}
\end{supptable}
